\documentclass[10pt,a4paper]{book}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\title{math & ml notes}
\begin{document}


\chapter{}
\begin{itemize}
	\item TensorArray
	
	Class wrapping dynamic-sized, per-time-step, write-once Tensor arrays.
\end{itemize}


\chapter{QA}
\section {Bazel}

\section {Protobuf}

\begin{itemize}
	\item How to generate debug version
	
	Add "--copt=-g" to bazel build options
	
	\begin{quote}
		bazel build --config=opt --copt=-g //tensorflow/tools/pip\_package:build\_pip\_package
	\end{quote}


	\item tensorflow's runtime architecture
	
	1)single machine
	2)Cluster
	

	\item what happens when a session is created?
	
	1) Search gpu device
	2) many threads are created
	3) Code: 
	  c api: TF\_NewSession, python

	Python class:
	
	SessionInterface 
	
	BaseSession 
	
	Session

	In init of BaseSession:
	
	1)get graph (default, or provided by user)
	
	2)A target maybe provided, what is it? An execution engine to connect to?
	
	3) call CAPI TF\_NewSession
	
	core/public/session.h
	core/common\_runtime/session.cc
	SessionFactory::GetFactory()->NewSession
	
	session factory manages many sessions, and they are compared to options of NewSession call, 
	if multiple factories are found, an error occures.
	
	Q: how many factories are there in tensorflow?

	\item What is a tensorflow execution engine?
	
	\item given a CAPI function name, what is the internal corresponding class/methods?
	
	find them in capi.cc
\end{itemize}


\end{document}
